---
output:
  github_document: default
---
```{r message=FALSE, warning=FALSE}
library(forecast)
library(ggplot2)
library(tidyr)
library(gridExtra)
```
# Autoreggresive Processes AR(p)
The Autoregressive process is given by the equation

$$X_t = \sum^p_{i=1} \phi_i X_{t-i} + w_t.$$

## Stationary AR(1)
AR(1) process
$$X_t = \phi X_{t-1} + w_t$$ 

is stationary if $|\phi| < 1$. In R we can simulate such process using `arima.sim` function. Greater absolute value results in greater "amplitude" and negative phi causes oscilations from positive to negative. See page 87 of https://sistemas.fciencias.unam.mx/~ediaz/Cursos/Estadistica3/Libros/Time%20Series%20Analysis%20and%20Its%20Applications.pdf. The proof of stationarity is based on the following:

$$X_t = \phi X_{t-1} + w_t$$
$$ X_t = \phi (\phi X_{t-2} + w_{t-1}) + w_t$$


$$ X_t = \phi (\phi (\phi X_{t-3} + w_{t-2}) + w_{t-1}) + w_t$$

$$X_t = \phi^3 X_{t-3} + \phi^2 w_{t-2} + \phi w_{t-1} + w_t$$

$$X_t = \phi^k X_{t-k} + \sum^k_{i=0} \phi^i w_{t-i}. \quad (eq.1)$$

```{r, fig.width = 18, fig.height = 5}
N <- 200
ar1_processes <- matrix(nrow = 7, ncol = N)
ar1_processes[1,] <- 1:N
ar1_processes[2,] <- arima.sim(model = list(ar = c(0.1)), n = N)
ar1_processes[3,] <- arima.sim(model = list(ar = c(0.5)), n = N)
ar1_processes[4,] <- arima.sim(model = list(ar = c(0.9)), n = N)
ar1_processes[5,] <- arima.sim(model = list(ar = c(-0.1)), n = N)
ar1_processes[6,] <- arima.sim(model = list(ar = c(-0.5)), n = N)
ar1_processes[7,] <- arima.sim(model = list(ar = c(-0.9)), n = N)

# Note: you can use arima.sim only to simulate a stationary process. To simulate non-stationary one,
# use Arima.

ar1_processes <- as.data.frame(t(ar1_processes)) 
colnames(ar1_processes) <- c(
  "time", "phi_0.1", "phi_0.5", "phi_0.9", "phi_-0.1", "phi_-0.5", "phi_-0.9")

plot_different_AR1 <- ar1_processes %>% pivot_longer(cols = 2:7) %>% 
  ggplot(aes(x = time, y = value, col = name)) +
  geom_line() + 
  ggtitle("AR(1) processes with different coefficient (only stationary, i.e. |phi| < 1 allowed by arima.sim)") +
  theme(legend.position = "bottom")

plot_different_AR1
```


### ACF of stationary AR(1) process
It can be shown that the autocovariance of AR(1) process with $|\phi| < 1$ is given by the formula
$$\gamma(h) = \frac{\sigma^2 \phi^h}{1 - \phi^2}.$$

It can also be shown that for a stationary process, autocorrelation $\rho(h) = \frac{\gamma(h)}{\gamma(0)}$. The autocorrelation function for a stationary
AR(1) process is simply
$$\rho(h) = \phi^h.$$

```{r, fig.width= 12, fig.height = 4}
ar1_acfs <- list()
for (name in colnames(ar1_processes)[-1]){
  ar1_acfs[[name]] <- ar1_processes[name] %>% ggAcf() +
    ggtitle(paste("AR(1) with", name))
}

gridExtra::grid.arrange(
  grobs = ar1_acfs, nrow = 2, top = "ACF plots for AR(1) process with different phi")
```


```{r, fig.width= 12, fig.height = 4}
ar1_pacfs <- list()
for (name in colnames(ar1_processes)[-1]){
  ar1_pacfs[[name]] <- ar1_processes[name] %>% ggPacf() +
    ggtitle(paste("AR(1) with", name))
}

gridExtra::grid.arrange(
  grobs = ar1_pacfs, nrow = 2, top = "PACF plots for AR(1) process with different phi")
```

### How to generate AR(1) process
```{r, fig.width = 15, fig.height = 40}
generate_AR_1 <- function(phi, sigma, n){
  epsilons <- rnorm(1+n, sd = sigma)
  for (i in 2:(n+1)){
    epsilons[i] <- epsilons[i] + phi*epsilons[i-1]
  }
  return(epsilons[-1])
} 

N <- 200
nsim <- 300
nplots <- 20
AR_1_by_hand <- lapply(1:nsim, function(x){generate_AR_1(phi = 0.9, sigma = 1, n = N)})
AR_1_builtin <- lapply(1:nsim, function(x){arima.sim(model = list(ar = c(0.9)), n = N)})
AR1_simulations <- list()
for (i in 1:nplots){
  # left column
  Y <- AR_1_by_hand[[i]]
  AR1_simulations[[2*i-1]] <- ggplot(
      data.frame(y = Y, x = 1:N)) +
    geom_line(aes(x,y)) +
    ggtitle(paste("By hand. sigma = ", round(sd(Y), 2), "mean = ", round(mean(Y),2))) +
    geom_hline(yintercept = mean(Y), col = "red") +
    scale_y_continuous(breaks = seq(-8,8,2), limits = c(-8,8))
  
  # right column
  Y <- AR_1_builtin[[i]]
  AR1_simulations[[2*i]] <- 
    ggplot(data.frame(y = Y, x = 1:N)) +
    geom_line(aes(x,y)) +
    ggtitle(paste("Built-in. sigma = ", round(sd(Y), 2),"mean = ", round(mean(Y),2))) +
    geom_hline(yintercept = mean(Y), col = "red") +
    scale_y_continuous(breaks = seq(-8,8,2), limits = c(-8,8))
}

gridExtra::grid.arrange(
  grobs = AR1_simulations, 
  ncol = 2, top = "AR(1) processes phi = 0.9 simulation")
```

```{r}
par(mfrow = c(1,2))
AR_1_builtin %>% sapply(mean) %>% boxplot(main = "means, builtin method", ylim = c(-3,3))
AR_1_by_hand %>% sapply(mean) %>% boxplot(main = "means, by hand", ylim = c(-3,3))
```

```{r}
par(mfrow = c(1,2))
AR_1_builtin %>% sapply(sd) %>% boxplot(main = "sd, builtin method", ylim = c(1,3))
AR_1_by_hand %>% sapply(sd) %>% boxplot(main = "sd, by hand", ylim = c(1,3))
```

### How to generate AR(p)?
```{r}
generate_AR_p <- function(AR_coefs, sigma, n){
  # Y_t = sum(phi_i Y_t-i + epsilon_i)
  p <- length(AR_coefs)
  epsilons <- rnorm(n+p, sd = sigma)
  for (i in 1:(n)){
    Y_i <- AR_coefs %*% Y + rnorm(1, sd = sigma)
    Y <- c(Y, Y_i)
  }
  return(Y)
}

# AR(1)
X <- generate_AR(
  AR_coefs = c(0.5),
  sigma = 1,
  n = 200
)

plot(X)
```

## Random walk
AR(1) process with an unit root $\phi = 1$ is called a random walk. It's expected value is $E(X_t) = t\mu$, where $w_t \sim N(\mu, \sigma^2)$ and it's variance is $Var(X_t) = \sum^t_{i=1} Var(w_t) = t\sigma^2$, therefore it is not stationary.




