---
output:
  github_document: default
---

```{r message=FALSE, warning=FALSE}
library(forecast)
library(ggplot2)
library(tidyr)
library(gridExtra)
```

# Types of noise 
## 1.1.1. White noise
Variable $w_t$ is a white noise if it meets only three conditions: 
- $E(w_t) = 0$,
- $Var(w_t) = \sigma^2$,
- $cov(w_t, w_s) = 0$.

Note it doesn't have to be normally distributed (then we call it "gaussian noise") - but in many cases it is. 

```{r, fig.width= 8, fig.height = 5}
set.seed(0)
white_noise = rnorm(200)
tsdisplay(white_noise, plot.type = "histogram", points = FALSE)
```

```{r}
mean(white_noise)
var(white_noise)
```

```{r}
set.seed(0) # for seed = 7 there is an accidental autocorrelation
white_noise2 = runif(200, min = -1, max = 1)
tsdisplay(white_noise2, plot.type = "histogram", points = FALSE)
```

```{r}
mean(white_noise2)
var(white_noise2) # 1/3 is a theoretical value
```

## Autoreggresive Processes AR(p)
The Autoregressive process is given by the equation

$$X_t = \sum^p_{i=1} \phi_i X_{t-i} + w_t.$$

### AR(1)
AR(1) process
$$X_t = \phi X_{t-1} + w_t$$ 


is stationary if $|\phi| < 1$. In R we can simulate such process using `arima.sim` function. Greater absolute value results in greater "amplitude" and negative phi causes oscilations from positive to negative.

See page 87 of https://sistemas.fciencias.unam.mx/~ediaz/Cursos/Estadistica3/Libros/Time%20Series%20Analysis%20and%20Its%20Applications.pdf.

##### Proof of stationarity
$$X_t = \phi X_{t-1} + w_t$$
$$ X_t = \phi (\phi X_{t-2} + w_{t-1}) + w_t$$


$$ X_t = \phi (\phi (\phi X_{t-3} + w_{t-2}) + w_{t-1}) + w_t$$

$$X_t = \phi^3 X_{t-3} + \phi^2 w_{t-2} + \phi w_{t-1} + w_t$$

$$X_t = \phi^k X_{t-k} + \sum^k_{i=0} \phi^i w_{t-i}.$$

```{r, fig.width = 18, fig.height = 5}
N <- 200
ar1_processes <- matrix(nrow = 7, ncol = N)
ar1_processes[1,] <- 1:N
ar1_processes[2,] <- arima.sim(model = list(ar = c(0.1)), n = N)
ar1_processes[3,] <- arima.sim(model = list(ar = c(0.5)), n = N)
ar1_processes[4,] <- arima.sim(model = list(ar = c(0.9)), n = N)
ar1_processes[5,] <- arima.sim(model = list(ar = c(-0.1)), n = N)
ar1_processes[6,] <- arima.sim(model = list(ar = c(-0.5)), n = N)
ar1_processes[7,] <- arima.sim(model = list(ar = c(-0.9)), n = N)

# Note: you can use arima.sim only to simulate a stationary process. To simulate non-stationary one,
# use Arima.

ar1_processes <- as.data.frame(t(ar1_processes)) 
colnames(ar1_processes) <- c(
  "time", "phi_0.1", "phi_0.5", "phi_0.9", "phi_-0.1", "phi_-0.5", "phi_-0.9")

ar1_processes %>% pivot_longer(cols = 2:7) %>% 
  ggplot(aes(x = time, y = value, col = name)) +
  geom_line() + 
  ggtitle("AR(1) processes with different coefficient (only stationary, i.e. |phi| < 1 allowed by arima.sim)") +
  theme(legend.position = "bottom")
```


```{r, fig.width= 12, fig.height = 4}
ar1_acfs <- list()
for (name in colnames(ar1_processes)[-1]){
  ar1_acfs[[name]] <- ar1_processes[name] %>% ggAcf() +
    ggtitle(paste("AR(1) with", name))
}

gridExtra::grid.arrange(
  grobs = ar1_acfs, nrow = 2, top = "ACF plots for AR(1) process with different phi")
```


```{r, fig.width= 12, fig.height = 4}
ar1_pacfs <- list()
for (name in colnames(ar1_processes)[-1]){
  ar1_pacfs[[name]] <- ar1_processes[name] %>% ggPacf() +
    ggtitle(paste("AR(1) with", name))
}

gridExtra::grid.arrange(
  grobs = ar1_pacfs, nrow = 2, top = "PACF plots for AR(1) process with different phi")
```



#### Stationary AR(1)


#### Random walk
AR(1) process with an unit root $\phi = 1$ is called a random walk. It's expected value is $E(X_t) = t\mu$, where $w_t \sim N(\mu, \sigma^2)$ and it's variance is $Var(X_t) = \sum^t_{i=1} Var(w_t) = t\sigma^2$, therefore it is not stationary.

```{r}
# generate_AR <- function(AR_coefs, sigma, n){
#   p <- length(AR_coefs)
#   Y <- rnorm(p, sd = sigma)
#   for (i in 1:(n-p)){
#     Y_i <- AR_coefs %*% Y + epsilon_i
#     Y <- c(Y, Y_i)
#   }
#   return(Y)
# }
# 
# # AR(1)
# X <- generate_AR(
#   AR_coefs = c(0.5),
#   sigma = 1,
#   n = 200
# )
```


